### 2010-07-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: explain the coding of Hutton's razor on the Indexed part
Reviewer 2 said:

"p9, fig. 7: A reminder near ExprID that the order of the sublists of
IndexedD needs to line up with the definition of Ty would be helpful."
Desc@ICFP: use dotted arrow for (i : I) -> X i -> Y i
Reviewer 2 said:

"p10.  In the types of applyI and substI, maybe you could employ an
abbreviation

IFun X Y = (i:I) -> X i -> Y i
IFix R X =  mu_I \overline{R^*_I X}

Then the type of, say, substI becomes the much more readable:

(I : Set)(R : TagDesc I)(X,Y : I -> Set) ->
(IFun X (IFix R Y)) -> (IFun (IFix R X) (IFi

In our universe of descriptions, the syntax of this language is
described by a tagged indexed description. We use the index to carry
the type: the resulting description is indexed by $\Ty$. We observe
that some constructors are always defined, namely $\SYMBEcond$
and $\SYMBEval$. On the other hand, the $\SYMBEplus$ and
$\SYMBEle$ constructors are index-dependent. $\SYMBEplus$ is
defined if and only if the result type -- the index -- is $\Enat$,
whereas $\SYMBEle$ is defined if and only if the index is
$\Ebool$. The actual code precisely follows this intuition, as shown
in Figure~\ref{fig:hexpr-full}.

@@

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. This corresponds to the
following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{G} \To
               \PITEL{\V{ty}}{\Ty} \\
             & (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
               \IMu{\Ty}{\HExprD}{\V{ty}} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumLeft \V{x})  & 
    \Con{\pair{\SYMBEval}{\V{x}}{}} \\
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumRight \V{v}) &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]

@@

We are now left with applying $\discharge$ over all variables of the
term. The type of this operation is the following:
%
\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{ty}}{\Ty}
             \PI{\V{G}}{\Ctxt} \\
           & (\PI{\V{ty}}{\Ty}
              (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
              \IMu{\Ty}{\HExprD}{\V{ty}}) \\
          & \To \IMu{\Ty}{\HExprVarD{\V{G}}}{\V{ty}} \To 
            \IMu{\Ty}{\HExprD}{\V{ty}}
\end{array}
\]

@@

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its
definition is the following:
%
\[\stk{
\begin{array}{@{}ll}
\substI : & _{\PI{\V{I}}{\Set}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
          & (\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
                               \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}) \To \\
          & \PI{\V{i}}{\V{I}}
            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{\V{i}} \To
            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
\end{array} \\
\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} = \\
\qquad    \cataI\: \toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}}\:
                      (\IMu{}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
                      \V{i}\:
                      \V{t} 
}\]
% 
Where  $\F{applyI}$ is defined as follow:
%
\[\stk{
\begin{array}{@{}ll}
\F{applyI} : & _{\PI{\V{I}}{\Set}}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
          & (\PI{\V{i}}{\V{I}} \V{X}\: \V{i} \To \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}) \To \\
          & \PI{\V{i}}{\V{I}} 
            \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                         {\V{I}}
                                         {\V{X}}}\: 
                     \V{i}}
                    {\V{I}}
{\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \To 
           \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
\end{array}
}\]

@@

The evaluator for closed terms we implemented in
Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}\: \V{ty}$ to values in
$\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
have to fill in the right arguments to $\substI$, the type guiding us:
%
\[\stk{
\substH\: \V{ty}\:
          \V{G}\:
          \V{g}\:
          \V{\sigma}\: 
          \V{term} \mapsto  \\
\quad \substI_{\Ty}\:
               (\SUM{\SYMBVal}{\SYMBVar\: \V{G}})\: 
               \SYMBVal\:
               \HExprFreeD\: 
               \V{\sigma}\:
               \V{ty}\:
               \V{term}
}\]
%
Hence completing our implementation of the open terms interpreter.

@@

In this section, we have presented the universe of indexed
description. It embraces indexed families of types and, as such,
allows us to write dependent datatypes. Hence, we have presented
several example of indexed datatypes. In this context, we have
presented the free monad construction, together with its monadic
operations.

>>>>

In our universe of descriptions, the syntax of this language is
described by a tagged indexed description. We use the index to carry
the type: the resulting description is indexed by $\Ty$. We observe
that some constructors are always defined, namely $\SYMBEcond$ and
$\SYMBEval$. On the other hand, the $\SYMBEplus$ and $\SYMBEle$
constructors are index-dependent. $\SYMBEplus$ is defined if and only
if the result type -- the index -- is $\Enat$, whereas $\SYMBEle$ is
defined if and only if the index is $\Ebool$. The actual code
precisely follows this intuition, as shown in
Figure~\ref{fig:hexpr-full}. The first element of $\HExprID$
corresponds to this choice of constructor: the sublists corresponds to
the constructors enabled by the corresponding type in $\Ty$.

@@

\newcommand{\DotTo}{\mathop{\blue{\dot{\rightarrow}}}}

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. The pointwise indexed
function type $\PI{\V{i}}{I}{X\: i \To Y\: i}$ is abbreviated  $X
\DotTo Y$. This corresponds to the following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{G} \To \\
             & (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
               \IMu{\Ty}{\HExprD}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumLeft \V{x})  & 
    \Con{\pair{\SYMBEval}{\V{x}}{}} \\
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumRight \V{v}) &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]

@@

We are now left with applying $\discharge$ over all variables of the
term. The type of this operation is the following:
%
\[
\begin{array}{@{}ll}
\substH  : & \PI{\V{G}}{\Ctxt} \\
           & ((\SUM{\Val{}}{\Var{\V{G}}{}}) \DotTo
              \IMu{\Ty}{\HExprD}{}) \DotTo \\
          & \IMu{\Ty}{\HExprVarD{\V{G}}}{} \DotTo 
            \IMu{\Ty}{\HExprD}{}
\end{array}
\]

@@

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its
definition is the following:
%
\[\stk{
\begin{array}{@{}ll}
\substI : & _{\PI{\V{I}}{\Set}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
          & (\V{X} \DotTo 
             \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To 
            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{} \DotTo
            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
\end{array} \\
\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} = \\
\qquad    \cataI\: \toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}}\:
                      (\IMu{}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
                      \V{i}\:
                      \V{t} 
}\]
% 
Where  $\F{applyI}$ is defined as follow:
%
\[\stk{
\begin{array}{@{}ll}
\F{applyI} : & _{\PI{\V{I}}{\Set}}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
          & (\V{X} \DotTo \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                         {\V{I}}
                                         {\V{X}}}}
                    {\V{I}}
                    {\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
\end{array}
}\]

@@

The evaluator for closed terms we implemented in
Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}\: \V{ty}$ to values in
$\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
have to fill in the right arguments to $\substI$, the type guiding us:
%
\[\stk{
\substH\: \V{G}\:
          \V{ty}\:          
          \V{g}\:
          \V{\sigma}\: 
          \V{term} \mapsto  \\
\quad \substI_{\Ty}\:
               (\SUM{\SYMBVal}{\SYMBVar\: \V{G}})\: 
               \SYMBVal\:
               \HExprFreeD\: 
               \V{\sigma}\:
               \V{ty}\:
               \V{term}
}\]
%
Hence completing our implementation of the open terms interpreter.

@@

In this section, we have presented the universe of indexed
description. It embraces indexed families of types and, as such,
allows us to write dependent datatypes. Hence, we have presented
several example of indexed datatypes. In this context, we have
presented the free monad construction, together with its monadic
operations. A summary of the components defining descriptions and
their indexed counterparts is presented in
Table~\ref{tab:sumup-operators}.

\begin{table}

{\small

\begin{tabular}{|c|c|c|c|}
\hline
Desc   & IDesc & Role & Status \\
\hline
\hline
\multicolumn{2}{|c|}{$\EnumU$}                  & Build finite sets           & Levitated \\
\hline
$\Desc$               & $\IDesc$               & Describe functors   & Levitated \\
\hline
$\SYMBdescop{\_}$         & $\SYMBidescop{\_}{\_}$         & Interpret descriptions      & Hardwired \\
\hline
$\SYMBMu$, $\SYMBCon$ & $\SYMBIMu_I$, $\SYMBCon$ & Define fixpoints & Hardwired \\
\hline
$\SYMBind$, $\SYMBAll$, $\SYMBall$ & $\inductionI$, $\SYMBAllI$, $\SYMBallI$ & Induction principle & Hardwired \\
\hline
\end{tabular}

}

\caption{Summary of constructions on Descriptions}
\label{tab:sumup-operators}

\end{table}
